--- 
title: "Statistics, R, and Music"
author: "David John Baker"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "This is a very early stage repo dedicated to creating a handbook for researchers of any level interested in learning more about statistics and the R language. Material from the book stems primarily from Jason Hicks' 4111 Intermediate Statistics class at LSU, as well as other sources I have encountered the past few years."
---

# Introduction to Statistics

What this book will assume 

* Basic knowledge of math
* Basic computer literacy
* Include chapter on levels of measuremet? Aka Chapter 1-4 Cohen?
* Ability to look for answers self, working with R you need to be able to troubleshoot

What book plans to accomplish

* Grounding in basic NHST statistics
* Teach basic calculations by hand
* Working knowldge of R for all tests
* Manipulating data in R
* Basic visualization of data in R
* Guide to finding similar values from SPSS

This is a _sample_ book written in **Markdown**. You can use anything that Pandoc's Markdown supports, e.g., a math equation $a^2 + b^2 = c^2$.

For now, you have to install the development versions of **bookdown** from Github:

```{r eval=FALSE}
devtools::install_github("rstudio/bookdown")
```

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.

To compile this example to PDF, you need to install XeLaTeX.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
